---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    fig_caption: yes
    highlight: pygments
    keep_md: yes
    theme: readable
---
## Overview

This analysis will evaluate data collected from a personal monitoring device.  The data tracks an anonymous subject's steps taken by day in five minute increments.  There are two months worth of data collected from October to November 2012.

## Loading and preprocessing the data

The following code will load the libraries I will use and then load the data into the dataframe d1. Then, I'll create a subset of d1 that excludes records with NA values called d2.

```{r, echo=T}

library(dplyr)
library(ggplot2)
library(DMwR)

d1<-read.csv(unz("activity.zip", "activity.csv"))

good<-complete.cases(d1)

d2<-d1[good,]
```

## What is mean total number of steps taken per day?

This first block of code will create a data frame (d3) that summarized the steps taken by date.

```{r, echo=T}

d3<-d2 %>%
  group_by(date) %>%
  summarize(steps=sum(steps))

```

This second block of code will plot a histogram of steps by day using the summarized data frame (d3).

```{r, echo=T}

plot1<-ggplot(data=d3, aes(d3$steps)) + geom_histogram(binwidth = 2500, col="black", fill="blue")
plot1<-plot1+labs(x="Steps", y="Count of Days", title="Histogram of Steps per Day")

plot1

```

Now, I will calculate and displsy the mean and median steps by day


```{r, echo=T}

options(scipen = 999)

stepmean<-round(mean(d3$steps),0)

stepmedian<-median(d3$steps)

```

The mean of steps taken is `r stepmean`

The median of steps taken is `r stepmedian`


## What is the average daily activity pattern?

This block of code will create another new data frame (d4) that summarizes steps taken by interval.

```{r, echo=T}

d4<-d2 %>%
  group_by(interval) %>%
  summarize(steps=mean(steps))

```

Next, I will plot the average steps by interval.

```{r, echo=T}

plot2<-ggplot(d4, aes(x=interval, y=steps))+geom_line(col="blue")+
  labs(x="Interval", y="Average Steps Taken", title="Average Steps Taken by Interval")

plot2

```

Now, we will find the interval with most steps taken

```{r, echo=T}

max<-max(d4$steps)

maxint<-subset(d4, steps==max)

```

The interval with the most steps taken is interval #`r maxint$interval`, which is equivalent to 8:35am.

## Imputing missing values

First we will calculate how many records in the original data frame (d1) have missing values.
 
```{r, echo=T}

NAnumb<-nrow(d1)-nrow(d2)

```

There are `r NAnumb` records with missing values

Next, I'll use the knnImputation() function from the DMwR package to impute the missing values.  I'll then summarize the imputed data by date.

```{r, echo=T}

d6<-knnImputation(d1, k=5)

d7<-d6 %>%
  group_by(date) %>%
  summarize(steps=sum(steps))

```

Now, I'll plot a histogram of steps by day using the imputed data

```{r, echo=T}

plot3<-ggplot(data=d7, aes(d7$steps)) + geom_histogram(binwidth = 2500, col="black", fill="blue")
plot3<-plot1+labs(x="Steps", y="Count of Days", title="Histogram of Steps per Day - Imputed N/A")

plot3

```

We can see that this histogram looks the same as the one generated with the N/A values excluded.

Now we'll see how much the imputing changed the mean and median number of steps taken by day

```{r, echo=T}

impstepmean<-round(mean(d7$steps),0)

impstepmedian<-median(d7$steps)

```

Imputed mean = `r impstepmean`

Imputing the missing data changes the average daily steps by `r round(impstepmean-stepmean, 0)`

Imputed median = `r impstepmedian`

Imputing the missing data changes the median by `r impstepmedian-stepmedian`


## Are there differences in activity patterns between weekdays and weekends?

First, I'll create a new factor variable called DayType that will indicate whether a given date is a weekend day or a weekday.  Then we will summarize the data by Interval and DayType.

```{r, echo=T}

d6$date<-as.Date(d6$date)

d6$weekday<-weekdays(d6$date)

d6<-mutate(d6, DayType = ifelse(test = weekday=="Saturday" | weekday=="Sunday", yes="Weekend", no="Weekday"))

d6$DayType<-as.factor(d6$DayType)

d8<-d6 %>%
  group_by(DayType, interval) %>%
  summarize(steps=mean(steps))

```

Now we'll produce a panel plot to compare step activity on weedays vs. weekends

```{r, echo=T}

plot4<-ggplot(d8, aes(x=interval, y=steps))+geom_line(col="blue")+
  labs(x="Interval", y="Average Steps Taken", title="Average Steps Taken by Interval: Weekday vs. Weekend")+
  facet_grid(DayType~.)

plot4

```

We can see from the plot that step activity was more constant throughout the day on weekends vs. weekdays.

